---
title: "Using DO.utils with reticulate and Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using DO.utils with reticulate and Python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(DO.utils)
```


# What does Python add?

DO.utils is designed to balance functionality and ease of use, with a goal for reproducibility. Some functionality is just not available in R, or it's less reproducible or harder to manage. Python fills in these gaps. Specifically using Python powers functions that:

1. Execute SPARQL queries on local files.
    - The R SPARQL engines require database management (`virtuoso`) or do not work with SPARQL 1.1 (`rdflib`).
2. Extract information from git repos.
    - To update data from releases and analyze unexpected changes to files.
3. Predict term mappings.
    - Utilizing the biopragmatics stack: bioregistry, biomappings, etc.; pyobo; and INDRA labs' GILDA


# Why use Python via reticulate?

Generally, the R-python connection allows data integration with R where data frames can be more easily managed, explored, and analyzed. More specifically, `reticulate` _should_ help manage Python and python module dependencies and simplify user setup/use.


# How to use python-dependent DO.utils functions?

These functions should work as standard R functions. In some cases internal data extractor functions have been written to eliminate the need for non-standard user data munging.


## Python Setup

Although it has not been tested yet, use of `reticulate` _should_ handle installation of python and/or python modules automatically and require little to no user input. The default is to attempt to identify and use python that is already installed. Should that not work, `reticulate` _should_ install python and needed modules via miniconda.


### Customization

More user control can be achieved over this installation (like using pip and venv instead of miniconda) by using reticulate to install python and create a virtual environment before calling `library(DO.utils)` for the first time. The basic procedure would be to execute the following:

```{r eval = FALSE}
library(reticulate) # installs with DO.utils
py_version <- "3.9" # or whatever version you prefer
venv_name <- "r-reticulate"

py_path <- install_python(py_version)
virtualenv_create(venv_name, py_path, version = py_version)
```


**A Note on Virtual Environments**

`reticulate` can create conda environments or venv/virtualenv virtual environments. Non-conda virtual environment options are described briefly below but similar options approaches are available for conda environments via `reticulate`.

`r-reticulate` is the default virtual environment name promoted by the developers of reticulate for _ALL_ packages that depend on `reticulate`, with the goals of 1) promoting interoperability between packages that depend on `reticulate` and python and 2) to better manage python module dependencies, through the use of a single virtual environment. Alternatives to this default environment can be specified by modifying the first argument `envname` to `virtualenv_create()`. A name (without slashes) will create a virtual environment of that name in the default location (see `virtualenv_root()`); a relative or absolute path (with slashes) will create a virtual environment in the specified directory, which can be nice if you want a python virtual environment associated with a specific project.


**A Note on Python Module Installation**

It is possible to specify python modules to install at the time a virtual environment is created or later by activating the virtual environment and executing `virtualenv_install()`. If only DO.utils dependencies are desired, this should not be necessary as the first execution of `library(DO.utils)` in an active virtual environment should install the python module dependencies of `DO.utils` automatically. Should any python module need to be installed by a user, the modules used by `DO.utils` can be found in the "Config/reticulate" section of it's `DESCRIPTION` file.



