---
title: "Using DO.utils with reticulate and Python"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{use-reticulate-python}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DO.utils)
```


# What does Python add?

DO.utils is designed to balance functionality and ease of use, with a goal for reproducibility. Some functionality is just not available in R, or it's less reproducible or harder to manage. Python fills in these gaps. Specifically using Python provides:

- A SPARQL engine (`rdflib`) that works with files, unlike R's `rdflib` which does not like `NOT` statements, and does not require managing a database (unlike the R `virtuoso` package).
- Access to Python-specific modules:
    - biopragmatics stack: pyobo, bioregistry, biomappings, etc.
    - INDRA labs' GILDA
    - _etc._
- Flexibility


# Why use Python via reticulate?

To keep everything in one place, minimize frustration, and maximize productivity.
