---
title: "How parse_omim_name() Works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How parse_omim_name() Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(DO.utils)
```

`parse_omim_name()` converts OMIM entry names from their raw all-uppercase,
inverted-filing format into normalized lowercase names with proper noun
capitalization. This article explains how the algorithm works, when
rearrangement is and isn't applied, what capitalization fixes are automatic,
and what the current limitations are.

---

## OMIM's inverted filing convention

OMIM lists entry names with the primary disease term first, followed by
comma-separated qualifier terms that in natural language would precede it:

```
SPASTIC PARAPLEGIA 14, AUTOSOMAL RECESSIVE
```

The natural form is: *autosomal recessive spastic paraplegia 14*.
`parse_omim_name()` reverses this inversion when recognized qualifiers are
present.

---

## Qualifier classification

Comma-separated tokens after the primary term are classified as follows:

### Pre-qualifiers — moved *before* the primary term

Adjective tokens such as `AUTOSOMAL RECESSIVE`, `CONGENITAL`, or `FAMILIAL`
are placed before the primary term. When multiple pre-qualifiers are present
they are reversed in order, so the last listed appears first — matching OMIM's
preferred-name conventions. Definitive inheritance terms (`AUTOSOMAL DOMINANT`,
`X-LINKED`, `Y-LINKED`, `MITOCHONDRIAL`) always take priority and appear first,
even when listed after other qualifiers.

```{r}
parse_omim_name("SPASTIC PARAPLEGIA 14, AUTOSOMAL RECESSIVE; SPG14")
parse_omim_name("DEAFNESS, AUTOSOMAL DOMINANT, CONGENITAL; DFNA")
```

### Post-qualifiers — attached *after* the primary term

* **Numeric & alphanumeric type codes** (`1`, `23`, `7A`) are appended with a space.
* **`TYPE ...` tokens** are appended with a space.

```{r}
parse_omim_name("OSTEOGENESIS IMPERFECTA, TYPE XI; OI11")
parse_omim_name("SCOLIOSIS, ISOLATED, SUSCEPTIBILITY TO, 1; IS1")
```

### Trailing phrases — appended *last*

Tokens beginning with `WITH`, `DUE TO`, `AND`, or `OR` are treated as
descriptive trailing phrases and appended after the primary term (and any
post-qualifiers).

```{r}
parse_omim_name("EPILEPSY, PROGRESSIVE MYOCLONIC, 4, WITH OR WITHOUT RENAL FAILURE; EPM4")
```

### `SUSCEPTIBILITY TO`

The token `SUSCEPTIBILITY TO` is always prepended before the entire assembled
name.

```{r}
parse_omim_name("DIABETES MELLITUS, SUSCEPTIBILITY TO")
```

---

## When rearrangement is applied

Rearrangement is only triggered when at least one qualifier matches a
**forcing pattern**:

* A pure number or alphanumeric subtype code (e.g. `1`, `7A`)
* A `TYPE` or `MULTIPLE TYPES` qualifier
* A definitive inheritance term (`AUTOSOMAL DOMINANT`, `X-LINKED`, etc.)
* A core set of strong adjective/onset qualifiers: `BILATERAL`,
  `CHILDHOOD-ONSET`, `CONGENITAL`, `EARLY-ONSET`, `FAMILIAL`, `FOCAL`,
  `GENERALIZED`, `HEREDITARY`, `HYPOMYELINATING`, `ISOLATED`, `JUVENILE`,
  `LATE-ONSET`, `NEONATAL`, `POSTSYNAPTIC`, `PRESYNAPTIC`, `PROGRESSIVE`,
  `SUSCEPTIBILITY TO`, `UNILATERAL`, `VESTIBULAR`

When **no** forcing pattern is present (e.g. multi-feature descriptor lists),
the tokens are kept in original comma order and only lowercased:

```{r}
# No forcing pattern → no rearrangement
parse_omim_name("SPASTIC TETRAPLEGIA, THIN CORPUS CALLOSUM, AND PROGRESSIVE MICROCEPHALY")
```

To extend the forcing list for edge cases, open an issue or submit a PR
editing `omim_has_forcing()` in `R/parse.R`.

---

## Capitalization fixes applied automatically

After rearrangement and lowercasing, `fix_disease_caps()` applies these rules:

* **Roman numerals** following `type ` are uppercased
  (e.g. `type xi` → `type XI`).
* **Alphanumeric subtype codes** are uppercased
  (e.g. `7a` → `7A`, `3a` → `3A`).
* **Immunoglobulin abbreviations**: `IgA`, `IgD`, `IgE`, `IgG`, `IgM`, `IgY`.
* **Word-level eponym substitutions** from the `eponyms` argument
  (default: [disease_eponyms]).
* **Phrase-level regex substitutions** from the `patterns` argument
  (default: [disease_cap_patterns]), applied longest-first and overriding
  any conflicting eponym substitutions.

---

## Known limitations

* **Unrecognized proper nouns** remain lowercase. Add missing eponyms to
  a custom `eponyms` vector, or submit a PR to extend [disease_eponyms].

* **Context-sensitive capitalization** (e.g. `SHORT` as an acronym vs `short`
  as an adjective) cannot be resolved by word-level `eponyms`. Use `patterns`
  with a phrase-level regex:

    ```{r}
    parse_omim_name(
        "SHORT SYNDROME; SHORTSYN",
        patterns = c("short syndrome" = "SHORT syndrome")
    )
    ```

* **Context-specific word substitutions** (e.g. `WITH` → `associated with`)
  cannot be inferred and require manual post-processing.

* **Qualifiers outside the forcing list** may not trigger rearrangement.
  The list covers the most common OMIM qualifier types; edge cases exist
  (submit an issue to extend it).
